\section{Process Abstraction}

A \textbf{process} abstracts the information required to describe and manage the execution of a program.

\subsection{Function Invocation}
Function invocation presents control flow and data storage challenges:
\begin{itemize}
    \item jumping to function body, and resuming after the function returns
    \item passing  parameters, and capturing a return value
    \item allocating local variables
\end{itemize}

\begin{defn}{stack memory}
    Stack memory is a region of memory separate from instruction and data memory that is used to store the following information for function invocations in \textbf{stack frames}:
    \begin{enumerate}
        \item return address (\code{PC}) of the caller
        \item arguments for the parameters of the function
        \item local variables used by the function
        \item saved \textbf{stack pointer (\code{SP})}
    \end{enumerate}

    Optionally, the stack frame also includes:
    \begin{enumerate}
        \item[5.] \textbf{frame pointer \code{(FP)}}
        \item[6.] saved registers
    \end{enumerate}
\end{defn}

Stack frames are \textbf{pushed} onto the stack on invocation and \textbf{popped} on return.

The \textbf{stack pointer \code{(SP)}} is a special register which contains the address of the first free location in the stack.

Because \code{SP} varies depending on the number of local variables, some processors provide a \textbf{frame pointer \code{(FP)}} which points to a fixed location in a stack frame.

When all general purpose registers are exhausted (\textbf{register spilling}), their values can be written to memory and restored at the end of the function call.

\begin{defn}{function invocation --- setup}
    If a function \code{f()} invokes \code{g()}, then \code{f()} is the \textbf{caller} and \code{g()} is the \textbf{callee}.

    \begin{itemize}
        \keyitem*{caller}{
            \begin{enumerate}
                \item passes arguments by setting registers directly and/or pushes arguments onto the stack
                \item saves the return address on the stack
            \end{enumerate}
        }
        \keyitem*{callee}{
            \begin{enumerate}
                \item saves the old stack pointer
                \item allocates space for local variables
                \item sets \code{SP} to the top of the stack
            \end{enumerate}
        }
    \end{itemize}
\end{defn}

\begin{defn*}{function invocation --- teardown}
    \begin{itemize}
        \keyitem*{callee}{
            \begin{enumerate}
                \item pushes return value onto the stack (if any)
                \item restores \code{SP} to the saved stack pointer
                \item sets \code{PC} to the return address
            \end{enumerate}
        }
        \keyitem*{caller}{
            \begin{enumerate}
                \item uses the return result (if any)
                \item resumes execution
            \end{enumerate}
        }
    \end{itemize}
\end{defn*}

Actual implementation of the function call convention depends on hardware, programming language, and compiler --- it is not universal.

\subsection{Dynamically Allocated Memory}
Some data may have a size only known at runtime, which precludes allocation in the data memory.

Some data may have an unknown lifetime (i.e. no deallocation time guarantees) which precludes allocation in the stack memory.

Hence, such dynamic data is allocated (e.g. by \code{malloc}) in the \textbf{heap memory}.

\subsection{Process Model}
\textbf{Process IDs} (PIDs) uniquely identify processes, each of which may exist in 5 different possible states:

\begin{enumerate}
    \keyitem*{new}{created, or under initialization}
    \keyitem*{ready}{admitted, but waiting to run}
    \keyitem*{running}{being executed by the CPU}
    \keyitem*{blocked}{waiting until an event occurs}
    \keyitem*{terminated}{finished execution, possibly requiring OS cleanup}
\end{enumerate}

\begin{defn}{process control blocks (PCBs)}
    Every process' registers, memory, PID, and state is stored in a \textbf{process control block}.
    
    The kernel manages all PCBs in a \textbf{process table}.
\end{defn}

\begin{defn}{context switching}
    Programs may be interrupted at any time, e.g. by the \textbf{scheduler},
    and must operate independently of interrupts.
    
    A \textbf{context switch} saves the current execution context process in a PCB and loads the context of the next process to be executed.
\end{defn}

\subsection{System Calls}
\textbf{System calls}, or syscalls, are an API for programs (or user-friendly libraries) to call services in the kernel.

Each syscall has a unique \textbf{system call number}, which is passed to the kernel via a designated register along with any arguments.

The signature of a syscall in C is \code{long syscall(long number[, arg1, arg2, ...])}.

These functions require a change from user mode to kernel mode when invoked, via a \textbf{TRAP instruction}.

Syscalls are laid out in a table, which a \textbf{dispatcher} uses the system call number to find the correct \textbf{system call handler}.

The system call handler executes the request, returns control to the library call, and switches back to user mode.

\subsection{Exceptions and Interrupts}
\textit{Machine level instructions} can cause \textbf{\textit{synchronous} exceptions} during program execution, e.g. division by zero or illegal memory access, which then passes control to an \textbf{exception handler}.

\textit{External events} such as timers or hardware can cause \textbf{\textit{asynchronous} interrupts}, which the operating then searches the \textbf{interrupt vector table} (IVT) to invokes the correct \textbf{interrupt handler}.

The IVT is set up by the OS during boot, and exists in memory (hardware).

As interrupts are asynchronous, they are only checked at specific times; never in the middle of an instruction: instructions are atomic/indivisible.

\begin{defn}{exception/interrupt handlers}
    These handlers save the register and CPU state, handle the necessary routine, and then restore the state and return to the interrupted program.

    Program execution may resume as though the exception or interrupt never occurred.
\end{defn}

\section{Unix Processes}
Every process has an integer process ID (PID).
Child processes also remember their parent's PID.

\code{init} is the ancestor of all other processes, created by the kernel at boot, with PID 1.

Process status can be viewed with the \code{ps} command.

\begin{defn}{\code{fork()}}
    \code{fork()} creates a new \textbf{child process} which shares the same code and address space as the parent process.

    Only the data memory (e.g. variables) of the child process is copied from the parent instead of shared.

    Because memory copying is expensive, some systems use \textbf{copy-on-write}.

    \code{fork()} returns \code{0} from the child process and the PID of the child process from the parent process.
\end{defn}

\begin{defn}{\code{execl()}}
    \code{execl()} replaces the current process with a new process specified by the first argument, with optional \code{NULL}-terminated command-line arguments:

    \begin{lstlisting}[frame=none,aboveskip=0.5em,belowskip=0em]
int execl(char *path, char *arg, ..., NULL);
    \end{lstlisting}
\end{defn}

\begin{defn}{wait()}
    \code{wait()} blocks the parent process until at least one child process terminates, after which it clears the remainder of the child process' resources which are not removed by \code{exit()}.

    \code{wait()} returns the PID of the terminated child process, while also accepting an optional \code{int} pointer for the child process' exit status:

    \begin{lstlisting}[frame=none,aboveskip=0.5em,belowskip=0.5em]
int wait(int *status);
    \end{lstlisting}

    If the parent process terminates before its child processes, the \textbf{orphan} child processes are adopted by \code{init} which will use \code{wait()} to cleanup.

    If the child process terminates with \code{exit()} but its parent does not call \code{wait()}, then it becomes a \textbf{zombie process}.
\end{defn}